---
title: "Introduction to Statistics"
author: "Week 1 codealong"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1

Load libraries: tidyverse and NHSRdatasets

```{r taks 1}
library(tidyverse)
library(NHSRdatasets)

```


## Task 2

Load and explore the LOS_model dataset

```{r task 2}
data("LOS_model")
?LOS_model
head(LOS_model, n=3)

# model1 <- glm(Death ~ Age + LOS, data=LOS_model, family="binomial")
# summary(model1)
```


## Task 3

Let's change the Death variable into a factor, so it has meaningful categories, with labels "Survived" and "Died."

```{r TASK 3}
LOS_model <- LOS_model %>% 
  mutate(Death = factor(as.character(Death), labels = c("Survived", "Died")))

LOS_model <- LOS_model %>% 
  mutate(Death = factor(as.character(Death), label = c("Survived", "Died")))
head(LOS_model)

```


## Task 4

Use the summary() function to calculate descriptive statistics for the dataset. Use the sd() function to calculate standard deviation.

```{r task 4}
summary(LOS_model)
sd(LOS_model$LOS)
sd(LOS_model$Age)
IQR(lOS_modedl$LOS)
```


## Task 5

Plot a histogram of Age in the sample.
```{r task 5}
LOS_model %>% 
  ggplot(aes(x = Age)) +
  geom_histogram()

LOS_model %>% 
  ggplot(aes(x = Age)) +
  geom_histogram(binwidth = 5, boundary = 5) +
  scale_x_continuous(breaks = seq(5, 100, by = 5)) +
  labs(title = "Distribution of Age in the LOS dataset")
```


## Task 6

Create a box plot of the age distribution in the data. 
```{r task 6}
LOS_model %>% 
  ggplot(aes(y = Age)) +
  geom_boxplot() +
  labs(title = "Distribution of Age in the LOS dataset")

LOS_model %>% 
  ggplot(aes(x = Organisation, y = Age)) +
  geom_boxplot() +
  labs(title = "Distribution of Age in the LOS dataset, by Trust")


```


## Task 7

Produce a bar chart of how many patients survived and died in each trust.
```{r task 7 } 
LOS_model %>% 
  ggplot(aes(x = Death)) +
  geom_bar()

LOS_model %>%
  ggplot(aes(x = Organisation, fill = Death)) +
  geom_bar(position = "stack")

LOS_model %>%
  ggplot(aes(x = Organisation, fill = Death)) +
  geom_bar(position = "stack") + 
  scale_fill_manual(values = c("gray", "red")) +
  labs(title = "Patient outcomes across NHS trusts",
      fill = "patient outcome" )
```


## Task 8

Produce a table of the number of patients in each trust who died in hospital.

```{r task 8}
LOS_model %>% 
  group_by(Organisation, Death) %>% 
  tally() %>% 
  filter(Death == "Died")
```


## Task 9

Let's create scatterplot of the relationship between length of stay in hospital and age. 

```{r task 9}
LOS_model %>% 
  ggplot(aes(x = Age, y = LOS, color = Death)) +
  geom_point() +
  labs(title = "Relationship between Age and LOS")


```

