---
title: "Cholera cases till 2023-07-31"
author: "write your name"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, load-packages}

pacman::p_load(
  data.table,             # fast to load large datasets
  rio,                    # to import/export different types of data
  here,                   # set relative path to project root
  
  dplyr,                  # data processing and manipulation
  lubridate,              # date manipulation
  Hmisc,                  # label variables
  janitor,                # data cleaning + quick tables
  epikit,                 # age categories
  stringr,                # working with string variables
  
  crosstable,             # generate tables
  gtsummary,              # generate tables
  flextable,              # enhance table visualization
 
  ggplot2,                # data visualization
  scales,
  incidence,              # epidemiological curve
  
  tidyverse               # data management
)

```

```{r, import-data}

cholera0 <-  import(here("Data", "Cholera", "cholera_20231102.csv"))

```

```{r, data-processing}

cholera <- cholera0 %>%     
  clean_names() %>% 
  rename(                  
    case_cat = f_diag
  ) %>% 
  mutate( sex= recode(sex,        
                      "f" = "female",
                      "F" = "female",
                      "Female" = "female",
                      "M" = "male",
                      "m" = "male",
                      "Male" = "male"
  ),
  age_yr = case_when(
    age_unit == "month" ~ round(age/12),
    .default = age
  ),
  age_cat = age_categories(age_yr,
                            lower = 0,
                            upper = 60,
                            by= 10)
  ) %>% 
  mutate(outcome_date= as.Date(outcome_date, origin= "1899-12-30"),
         adm_date= mdy(adm_date)
           ) %>%
  select(-c(muac, ocv_doses, starts_with(c("rdt_", "cult_"))))

```

# Executive summary

As of `r max(cholera$adm_date, na.rm= T)`, there were a total of **`r nrow(filter(cholera, case_cat== "confirmed"))`** confirmed cholera cases and **`r nrow(filter(cholera, case_cat== "confirmed" & outcome== "death"))`** deaths(CFR = **`r percent(nrow(filter(cholera, case_cat== "confirmed" & outcome== "death"))/nrow(filter(cholera, case_cat== "confirmed")))`**). The mean age of confirmed cases was **`r mean(cholera$age_yr[cholera$case_cat== "confirmed"], na.rm= T)`** years (standard deviation = **`r round(sd(cholera$age_yr[cholera$case_cat== "confirmed"], na.rm= T), digits=2)`**)



## Reported cholera cases by age categories, till 31 July 2023  

```{r, table1}
crosstable(cholera, c(age_cat)) %>% flextable::as_flextable()
```


## Distribution of reported cholera cases by date of admission and case category, till 31 July 2023  

```{r, graph1}

weeks <-  seq.Date(
  from = floor_date(min(cholera$adm_date, na.rm = T), unit = "week", week_start = 7),
  to = ceiling_date(max(cholera$adm_date, na.rm = T), unit = "week", week_start = 7),
  by = "week"
)

epicurve <- cholera %>% 
  ggplot(aes(x= adm_date, fill= case_cat))+
  geom_histogram(
    breaks = weeks,
    closed = "left",
    color = "grey50"
  )+
  scale_y_continuous(expand = c(0,0))+
  scale_x_date(
    expand = c(0,0),
    breaks = seq.Date(
      from = floor_date(min(cholera$adm_date, na.rm=T), unit = "week", week_start = 7),    
      to =   ceiling_date(max(cholera$adm_date, na.rm=T), unit = "week", week_start = 7),
      by =   "6 weeks"),
    
    minor_breaks = seq.Date(
      from = floor_date(min(cholera$adm_date, na.rm=T), unit = "week", week_start = 7),    
      to =   ceiling_date(max(cholera$adm_date, na.rm=T), unit = "week", week_start = 7),
      by =   "week"),
    
    date_labels = str_glue("%Y-W%U"))+
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1))
  
epicurve
```
