---
title: "Mpox situation report till 31 August 2023"
date: "2023-10-16"
output:
  word_document:
   reference_docx: "template.docx"
   toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-packages}
pacman::p_load(                    #load packages required for type of analysis in the report
  data.table,
  rio,
  here,
  
  janitor,
  dplyr,
  epikit,
  lubridate,
  aweek,
  
  scales,
  apyramid,
  crosstable,
  ggplot2,
  flextable,
  gtsummary,
  tidyverse
)

```

```{r import-data}

## we can use base R function read.csv 

# mpox <- read.csv(here("output", "mpox_20231012.csv")) # we used here function to create relative path to the project root

## another option is read_csv function from readr package

# mpox <- read_csv(here("output/mpox_20231012.csv")) 

## importing data with import/here functions

mpox <-  import(here("Data", "Mpox", "mpox_20231012.csv"))

## using fread function

# path <- "write here your directory path"   # do not forget to switch the \ to /
# 
# mpox <- fread(file = paste0(path, "write here your file name.csv"), header = TRUE)
```



```{r data-processing}
mpox <- mpox %>% 
 # we can create epiweek using epiweek function from lubridate
  dplyr::mutate(epiweek = epiweek(adm_date),     
 # another option is date2week function from aweek           
               epiweek2 =  aweek::date2week(    
                 adm_date,
                 week_start = "sun", 
                 floor_day = T),
  # start day of the week              
               epiweek_date = as.Date(epiweek2),
 # another option using lubridate function
               # epiweek_date2 = floor_date(adm_date,
               #                            unit = "week",
               #                            week_start = 7
               #                            ),
               year = year(adm_date),
               
               month = month(adm_date),
 # age grouping could be done with cut function from base R           
               age_gp = age_categories(  # check other arguments of age_categories function
                 age_yr,
                 breakers = c(0, 10, 20, 25, 30, 35, 40, 45, 55, 65),
                 separator = "-"),
 
               symptomatic= case_when(
                 symp_onset != is.na(symp_onset)  ~ "yes",
                 TRUE                             ~ "no"),
               
               case_outcome = case_when(
                 outcome == "death" ~ "died",
                 outcome == "unknown"|outcome == "" ~ "unknown",
                 .default = "alive"))

```

```{r filter-pos-cases}

mpox_pos <- mpox %>% 
  filter(result == "positive")
```

\pagebreak

# Summary  
\n  
\n  


As of `r max(mpox$adm_date, na.rm = T)`, a total of **`r nrow(mpox)`** suspect monkeypox cases were reported from **`r length(unique(mpox$country))`**. Out of _`r nrow(mpox)`_ reported cases, **`r fmt_count(mpox, result == "positive")`** confirmed mpox cases by PCR test. \n

[**_From positive cases_**]{.ul}, `r fmt_count(mpox_pos, case_outcome == "died")` fatal cases were recorded. Males represented `r fmt_count(mpox_pos, sex == "male")` of total positive cases. The median(IQR) age of positive cases was `r median(mpox_pos$age_yr, na.rm = T)` (`r quantile(mpox_pos$age_yr, prob = c(0.25, 0.75))`). There are `r mpox_pos %>% filter(year == 2022) %>% nrow()` and `r mpox_pos %>% filter(year == 2023) %>% nrow()` positive cases in 2022 and 2023, respectively.

\pagebreak

# **1. Person**

**Table 1: Mpox confirmed cases by age group and sex, till 31 July 2023**

```{r table1}

## tabyl and adorn

#  epiweek_table <- mpox %>% 
#   filter(result == "positive", sex != "intersex") %>% 
#   tabyl(age_gp, sex, show_na = F) %>%   # note the default is show_na = TRUE
#   adorn_totals(where = "both") %>%
#   adorn_percentages("col") %>%          # we can arrange using arrange(desc(name of col))
#   adorn_pct_formatting(digits = 0) %>% 
#   adorn_ns(position = "front") %>% 
#   flextable::as_flextable()
# 
# epiweek_table


## tidyverse option 

# mpox %>% 
#   filter(result == "positive", sex != "intersex") %>% 
#   group_by(age_gp) %>% 
#   summarise(
#     female = sum(sex == "female", na.rm = T),
#     male = sum(sex == "male", na.rm = T),
#      total = n()) %>% 
#   adorn_percentages("col") %>% 
#   adorn_pct_formatting(digits = 0) %>% 
#   adorn_ns(position = "front") %>% 
#   flextable::flextable() %>% 
#   flextable::autofit()


# ## crosstable
# 
# crosstable(mpox_pos %>% filter(sex != "intersex"), 
#            cols = c(age_gp), total = "col",
#            by = c(sex), percent_pattern = "{n} ({p_row})", 
#            percent_digits = 1) %>% 
# flextable::as_flextable()


## gt summary

mpox %>% 
  filter(result == "positive", sex != "intersex") %>%
  dplyr::select(age_gp, sex) %>% 
  tbl_summary(by = sex, label = list(age_gp ~ "Age group")) 
  
```

\pagebreak


***Figure(1): Age-sex pyramid of reported Mpox cases till 31 August 2023***

```{r fig1}

mpox %>% 
  filter(sex != "intersex") %>% 
  apyramid::age_pyramid(
    age_group = age_gp,
    split_by = sex,
    pal = c("lightblue", "pink"),
    proportional = T)+
  labs(x = "Age group (year)",
       title = "Age and sex distribution of admitted mpox cases",
       subtitle = "Data till 31 August 2023",
       caption = "Note: admitted cases include suspect, propable, and confirmed cases \n Mock data was used to create this figure")+
  theme(axis.text = element_text(face = "bold"),
        axis.title = element_text(face = "bold", color = "darkblue"),
         legend.title = element_blank(),
        plot.caption.position = "plot") +
  scale_fill_manual(
    values =  c("lightblue", "pink"),
    labels = c("male" = "Male", "female" = "Female")
  ) 

```

\pagebreak

***Figure(2): Distribution of confirmed Mpox cases' outcome by symptom presence***

```{r fig2}

# barplot

  mpox_pos %>% 
    ggplot(aes(x= case_outcome, fill = symptomatic))+
  geom_bar(position = "dodge") +
  coord_flip()
```

\pagebreak

# **2. Place**

**Table 2: Confirmed Mpox cases by country**\
*Date as of 31 August 2023*

```{r table2}

table2 <- mpox_pos %>% 
  group_by(country) %>% 
  summarise(
    'Total cases' = n(),
    'Total death' = sum(outcome == "death", na.rm = TRUE),
    cases_Jul2023 = sum(adm_date >= "2023-07-01" & adm_date <= "2023-07-31"),
    cases_Aug2023 = sum(adm_date >= "2023-08-01" & adm_date <= "2023-08-31"),
    Monthly_change = percent((cases_Aug2023 - cases_Jul2023)/cases_Jul2023),
    'Most recent admission date' = max(adm_date, na.rm = T)
  ) %>% 
  flextable() %>%               # we may go further with table beautification using flextable options
   flextable::set_header_labels(country = "Country",
                               `cases_Jul2023` = "Cases in July 2023",
                               cases_Aug2023 = "Cases in August 2023",
                               Monthly_change = "Monthly change %") %>% 
  flextable::bold(part = "header") %>% 
  flextable::bold(j =1) %>% 
  flextable::theme_zebra() %>% 
  flextable::bg(part = "header",
               bg= "#5DADE2") %>% 
  flextable::bg( j =6,
                 i= ~  Monthly_change >= 10,
                 bg= "red",
                 part= "body"
               )  %>%
  flextable::bg( j =6,
                 i= ~  Monthly_change < 10,
                 bg= "lightgreen",
                 part= "body"
               ) %>% 
  flextable::color(part = "header",
                   color = "white") %>% 
  flextable:: align(part = "header", j=2:3, align = "center") %>% 
  flextable::align(part = "body",j=2:3, align = "center") %>% 
  autofit()  %>% 
  set_table_properties(layout = "autofit")
table2
```


\pagebreak


# **3. Time**


***Figure(3): Weekly distribution of confirmed mpox cases by reporting country and date of admission***\
Date till 2023-08-31

```{r fig3}

# daily epi curve

# mpox %>% 
#   filter(result == "positive") %>% 
#   ggplot(aes(x= adm_date, fill = country))+
#   geom_histogram(
#     binwidth = 7,
#     color = "darkgrey"
#   )+
#   scale_x_date(
#     expand = c(0,0),
#     date_breaks = "6 weeks",
#     labels = scales :: label_date_short()
#   )+
#   scale_y_continuous( expand = c(0,0)) +
#   scale_fill_manual(
#     values = c("lightblue", "pink", "lightgreen", "lightyellow", "lightgrey")
#   )+
#   labs(
#     x= "Admission date",
#     y= "Number",
#     fill= ""
#   )+
#   theme(
#     axis.text = element_text(face = "bold", color = "black"),
#     axis.title = element_text(face = "bold", color = "black")
#   ) +
#   theme_classic()

```


```{r}

## Sunday as start day of the  week

weeks <- seq.Date(
  from = floor_date(min(mpox$adm_date, na.rm=T), unit = "week", week_start = 7),     
  to =   ceiling_date(max(mpox$adm_date, na.rm=T), unit = "week", week_start = 7),
  by =   "week")


mpox %>% 
  filter(result == "positive") %>% 
  ggplot(aes(x= adm_date, group = country, fill = country))+  
  geom_histogram(
    breaks = weeks,
    closed = "left",
    color = "darkgrey"
  )+
  scale_x_date(
    expand = c(0,0),
    
    breaks = seq.Date(
  from = floor_date(min(mpox$adm_date, na.rm=T), unit = "week", week_start = 7),    
  to =   ceiling_date(max(mpox$adm_date, na.rm=T), unit = "week", week_start = 7),
  by =   "6 weeks"),
  
  minor_breaks = seq.Date(
  from = floor_date(min(mpox$adm_date, na.rm=T), unit = "week", week_start = 7),    
  to =   ceiling_date(max(mpox$adm_date, na.rm=T), unit = "week", week_start = 7),
  by =   "week"),
  
  date_labels = str_glue("%Y-W%U")) + #  we can add \n after year, to enter week number to new line 
  
  scale_y_continuous( expand = c(0,0),
                      labels = label_number(accuracy = 1)) + # to round to whole number
  scale_fill_manual(  # to manual modification of colors or change legend labels or add legend title
    values = c("lightblue", "pink", "lightgreen", "lightyellow", "lightgrey")
  )+
  labs(
    x= "Admission date",
    y= "Number",
    fill= ""
  )+
  theme_classic()+
  theme(
    axis.text = element_text(face = "bold", color = "black"),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title = element_text(face = "bold", color = "black")
  )

```









